---
title: "Get Started"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{metrics-in-r}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## I'm totally new to R!

Welcome to the carnival 🎪. R is a great choice of programming language to learn for anyone interested in econometrics or data science more broadly. It's extremely popular, has high-quality tooling, and an incredibly inclusive community. Plus, it's free!

As mentioned in the home page, [R for Data Science](https://r4ds.had.co.nz/) is a great introductory text, and you should read it throughout the semester to gain new R skills. For the purposes of getting started with this website, I suggest at least reading the Introductory chapters as well as the "Workflow: basics" and "Workflow: scripts" chapters before returning to this page.

## Installation

This package is not hosted on CRAN, so you can't install it with the `install.packages` command. You need to use `devtools` to install it from my GitHub page:

```{r, eval = FALSE}
library(devtools)
install_github("nateybear/metrics-in-r")
```

You should see a bunch of other things installing too. This is the point: by installing my list of curated packages you should be "set up" and ready to start doing your homeworks in R.

## Resources

There are many great resources for learning R and specific aspects of R. Here is a non-exhaustive list that I will update occasionally:

* [R for Data Science](https://r4ds.had.co.nz/) by Hadley Wickham and Garrett Grolemund
  - Mentioned above
* [Introduction to Econometrics with R](https://www.econometrics-with-r.org/) by Christoph Hanck, Martin Arnold, Alexander Gerber, and Martin Schmelzer
  - This book mirrors Jeffery Wooldridge's Introduction to Econometrics and uses R to demonstrate how to do each of the examples. Extremely helpful if you are using Wooldridge for your class.
* [Advanced R](https://adv-r.hadley.nz/) by Hadley Wickham
  - What it says on the cover. If you already have some experience with R or programming in general, this is a deeper diver to learn how the language works under the hood.

