---
title: "Robust Standard Errors"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Robust Standard Errors}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

Congrats! You've managed to make a linear model in R, but the immediate problem that us economists face is that we want robust standard errors. R doesn't provide this out of the box, so we have to use a package that will compute robust standard errors for us. Luckily, this is quite easy to do, and we'll run through it here in this article.

## `sandwich` does it all

The `sandwich` R package is a Swiss army knife that can do a lot for you when it comes to robust standard errors. It's called "sandwich" because most variance estimators tend to have a form that looks like a sandwich: that's the $Q_{XX}Q_{u^2XX}Q_{XX}$ form you may be familiar with. Either way, I think it's a cute name.

Let's recreate the model that we used in the intro article.

```{r, message = F}
library(tidyverse)
library(haven)

htv <- read_dta(system.file("HTV.DTA", package = "metrics.in.r"))
model <- lm(wage ~ educ * abil, data = htv)
```

Now we load the sandwich package.

```{r}
library(sandwich)
```

There are a few different functions in sandwich, which correspond to the different types of robust variance estimators: `vcovHC` for the classic "White's" (heteroskedasticity-consistent) standard errors; `vcovCL` for clustered standard errors; and `vcovBS` for boostrapped standard errors. Making a covariance matrix of our $\beta$ estimates for a given regression model is as easy as calling `vcovHC` on the model. The `type` argument is optional: There are different types of "corrections" for heteroskedasticity-robust standard errors, and HC1 is the correction that Stata uses by default.

```{r}
vcovHC(model, type = "HC1")
```

From now on, I'll omit the `type` argument to keep my R code cleaner. Once we have the robust covariance matrix, we can get the robust standard errors of our $\beta$ estimates by taking the square root of the diagonal:

```{r}
sqrt(diag(vcovHC(model)))
```

Another way to do this is with the pipe operator:

```{r, eval=F}
model %>% vcovHC %>% diag %>% sqrt
```

I won't explain the pipe operator now. I prefer you to have an "immersive" introduction for now, and I will cover it more in-depth later. Just notice that it sort of "unwraps" the nested function calls and reverses the order of the calls.

## But what about the pretty tables

Recall that in the intro article we could do this with our model and get a nice summary table

```{r}
summary(model)
```

This summary table was generated with a homoskedastic assumption on the variance of the residuals, so we need to figure something else out for our robust standard errors. Luckily, there is a package called `lmtest` which provides this functionality. We call the `coeftest` function and give it both our model and a covariance matrix, and it will print a nice summary table like above:

```{r, message = F, warning=F}
library(lmtest)
```
```{r}
coeftest(model, vcov. = vcovHC(model))
```

Again, with the pipe this would look like

```{r, eval=F}
model %>% coeftest(vcov. = vcovHC(.))
```

So we have to expend a little more effort to get the covariance matrix of our $\beta$ estimates, but once we have both the model and our covariance matrix, we can make a summary table. In fact, we really could do this ourselves. I'm going to give a demonstration below that all we _need_ for these summary tables is a model and a covariance matrix. Let's replicate the output of `coeftest`—you don't need to understand what I'm doing, just notice that the only objects that I'm using are the model and the covariance matrix.

```{r}
estimate <- coef(model)
std_error <- model %>% vcovHC %>% diag %>% sqrt
t_value <- estimate / std_error
p_value <- 2 * pt(-abs(t_value), df = df.residual(model))

cbind(estimate, std_error, t_value, p_value)
```

## There are easier alternatives

Since this is R and there are a lot of people using it, it makes sense that someone has written something to make robust standard errors easier. Two people have in fact.

The first is the `lm_robust` function from the `estimatr` package. You specify the covariance type when you create your model, and then the `summary` function will use the correct type of robust standard errors in its output:

```{r}
library(estimatr)

model <- lm_robust(wage ~ educ * abil, data = htv, se_type = "HC1")

summary(model)
```

The second (and preferred) method is the `feols` function from the `fixest` package. Just like `lm_robust`, you specify the covariance type when you create your model:

```{r}
library(fixest)

model <- feols(wage ~ educ * abil, data = htv, vcov = "HC1")

summary(model)
```

**A word of warning**: There are issues computing clustered standard errors with `lm_robust`. In general, I don't trust the results that `lm_robust` give as much as I trust `sandwich` or `fixest`. Keep that in mind and choose wisely. I show you `lm_robust` because there are still many people that use it. You should convince them to switch over to `fixest` 🙂

## Conclusion: Why'd you show me the hard way?

Two reasons why I showed the `sandwich` and `coeftest` method. One is that I want to demystify a little bit of what is going on behind the scenes. You only need a _model_ and a _covariance matrix_. Once you have those, you can compute standard errors, t statistics, and p values. I prefer using `sandwich` exactly for this reason: I don't like "magic" and functions that do too much for me without me realizing what is going on.

The second reason is that the `sandwich` and `coeftest` method remains enduringly popular. `lm_robust` had a moment of fame, and `fixest` is up and coming as a very popular library in econometrics, but sometimes you just want a simple, tried-and-true method. You can choose for yourself which method you'd like to use.

<hr/>

Happy Coding!